---
title: "Coursera Reproducible Research Project 2"
author: "Mo Hao"
date: "21 May 2017"
output: md_document
---

# An Analysis of Impact from Severe Weather Events on Population Health and Economic Damange in U.S. from 1950 to Nov 2011

## Synopsis
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, and an analysis is subsequently done to study the impact from severe weather events in U.S. from year 1950 and end in November 2011. The raw data are loaded and processed in R. The total impact of population health and economic damage are aggregated by each disaster type and plotted in barcharts. In the result section, Tornado and Flood are found to be the most severe weather events in impacting population health and economic damage respectively. In the last section, the Tornado location is plotted on top of map to identify the area that are prone to Tornado disaster. 

## Data Processing
### Loading and reading the data
```{r}
# Setup the WD, download the file
setwd('C:/Users/skyqi/Desktop/Coursera_Reproducible Research_PA 2')
if(!file.exists('data')) dir.create('data')
fileUrl <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download.file(fileUrl,destfile='./data/repdata_data_StormData.csv.bz2')
# Load readr package to use read_csv function and load the data
library(readr)
StormData <- read_csv('./data/repdata_data_StormData.csv.bz2',col_types='dcccdcccdccccdcdccdddddddcdccccddddcd',progress=F)
```

### Setting up functions
1. setup function to sort data decendingly
```{r}
# function of sorting data decendingly
sortDF <- function(df, by){
  df <- df[order(by, decreasing = T),]
  return (df)
}
```

2. Setup function to transform EXP to integers
```{r}
# function of transforming EXP to integers
transformEXP <- function(var){
  if (class(var)!= 'character') var= sapply(var, as.character)
  x=tolower(var)
  x[x=='b']<-9
  x[x=='h']<-2
  x[x=='k']<-3
  x[x=='m']<-6
  x[x %in% c("","-","?","+")]<-0
  x=as.integer(x)
  return (x)
}
```

### Processing data
1. Aggregate Fatality and Injury by Type
```{r}
StormData$healthImpact<-StormData$FATALITIES+StormData$INJURIES
healthImpactByType<-aggregate(healthImpact~EVTYPE,data=StormData,sum)
healthImpactByType<-sortDF(healthImpactByType,healthImpactByType$healthImpact)
```

2. Aggregate Economic Loss by Type
```{r}
StormData$PROPDMGEXP<-transformEXP(StormData$PROPDMGEXP)
StormData$CROPDMGEXP<-transformEXP(StormData$CROPDMGEXP)
StormData$PropLoss<-with(StormData,(PROPDMG*(10**PROPDMGEXP)+CROPDMG*(10**CROPDMGEXP)))
PropLossByType<-aggregate(PropLoss~EVTYPE,data=StormData,sum)
PropLossByType<-sortDF(PropLossByType,PropLossByType$PropLoss)
```


## Results
### Plot the top 10 disaster types:
Tornado is found to be the most severe disaster in U.S. in terms of population health.   
Flood is found to be the most severe disaster in U.S. in terms of economic damage.    
```{r}
library(ggplot2)
ggplot(healthImpactByType[1:10,],aes(EVTYPE,healthImpact))+geom_bar(stat='identity',fill='red')+theme_bw()+coord_flip()+
  labs(title='Weather Impact on Health',x='Type of Weather Events',y='Fatality+Injury')
ggplot(PropLossByType[1:10,],aes(EVTYPE,PropLoss))+geom_bar(stat='identity',fill='blue')+theme_bw()+coord_flip()+
  labs(title='Weather Impact on Economic Loss',x='Type of Weather Events',y='Property Loss+Crop Loss')
```

### Plot the location of the Tornado
```{r}
library(ggmap)
Tornado<-StormData[StormData$EVTYPE=='TORNADO',]
Tornado$LONGITUDE<-(-Tornado$LONGITUDE/100)
Tornado$LATITUDE<-Tornado$LATITUDE/100
map<-get_map('United States',zoom=4)
Tornado$size<-round(Tornado$healthImpact/max(Tornado$healthImpact) *10 ,2)
ggmap(map)+geom_jitter(aes(LONGITUDE,LATITUDE,size=size),data=Tornado,alpha=0.1,colour='red')
```

The tornado is mostly happened in center and southeast coast of the United States. 